title: "Series-02-Rstudio : corelations and statistical test"
author: "Matys PRÉCLOUX"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    number_sections: true
---

# Correlations and linear regression
In this series of exercises, you will explore correlations and linear regression, using a dataset from a real experiment.

## The problem
You want to know if there is a relationship between the nonuse of the proximal part of the upper limb (PANU) and the nonuse of the shoulder (SANU) or the elbow (EENU).

## The data
As you will see, the data is far from perfect, but this is the reality of experimental data, especially in the case of clinical data.

Download the file NonUse.csv from the course Moodle.

```{r load data}
nonuse <- read.csv("data/test_data/NonUse.csv", sep=',', header=TRUE)
summary(nonuse)

```


This file contains measures of upper extremity nonuse:

    PANU: Proximal Arm Non Use
    SANU: Shoulder Antepulsion Non Use
    EENU: Elbow Extension Non Use

In your notebook, make a clear description of the experiment, the variables and the data. Do not forget to give the units of each variable, and the range of possible values.

| name | description | unit | range |
|--    |---------     |--   |---    |
| PANU | Proximal Arm Non Use         | ‰ | 0-100 |
| SANU | Shoulder Antepulsion Non Use | ‰ | 0-100 |
| EENU | Elbow Extension Non Use      | ‰ | 0-100 |



## Corelation analysis
Below are some questions to guide your analysis.

- Is PANU correlated with SANU and/or EENU?

we use spearman correlation bc we do not have assumptions about the distributions of the data, and we do not assume linear relationships between the variables 

```{r correlation}
cor.test(nonuse$PANU, nonuse$SANU, method = "spearman")

```


- Visualize the PANU-SANU relationship (and the PANU-EENU relationship)

```{r plot sanu+lm}
library(ggplot2)
ggplot(nonuse, aes(x=PANU, y=SANU)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, color='blue', se=T) + 
  #xlim(-100, 100) +
  #ylim(-100, 100) +
  labs(title= "PANU vs SANU", x="PANU(x)", y="SANU(%)") + 
  theme_minimal()

library(ggplot2)
ggplot(nonuse, aes(x=PANU, y=EENU)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, color='blue', se=T) + 
  #xlim(-100, 100) +
  #ylim(-100, 100) +
  labs(title= "PANU vs EENU", x="PANU(x)", y="EENU(%)") + 
  theme_minimal()

```


-linear regression for PANU-SANU (and for PANU-EENU)

```{r}
lm_sanu <- lm(SANU ~ PANU, data=nonuse)
summary(lm_sanu)
coef(lm_sanu)
slope <- coef(lm_sanu)[2]
intercept <- coef(lm_sanu)[1]
cat(" Regression equation : SANU=", slope, "*PANU +", intercept, "\n")
```
linear regression for PANU-EENU

```{r PANU-EENU}
lm_eenu <- lm(EENU ~ PANU, data=nonuse)
summary(lm_eenu)
coef(lm_eenu)
slope <- coef(lm_eenu)[2]
intercept <- coef(lm_eenu)[1]
cat(" Regression equation : EENU=", slope, "*PANU +", intercept, "\n")
```


- regression equations: PANU = a * SANU + b and PANU = a * EENU + b

```{r equations}
slope <- 0.08932839
intercept <- 0.5860521

a <- 1 / slope
b <- -intercept / slope

cat("Regression equation : PANU =", a, "* SANU +", b, "\n")

slope <- 0.5646221
intercept <- 3.283697
# PANU = a*EENU + b
a <- 1 / slope
b <- -intercept / slope

cat("Regression equation : PANU =", a, "* EENU +", b, "\n")
```


- Summarize results : 

Our correlation and regression analyses showed that PANU was more strongly associated with EENU than with SANU. The regression equation indicated that a one-unit increase in EENU was associated with an estimated 1.77-unit increase in PANU (PANU = 1.77·EENU – 5.82). By contrast, the relationship with SANU was weaker, with the regression equation suggesting that a one-unit increase in SANU corresponded to an estimated 11.20-unit increase in PANU, but with a higher intercept and greater variability (PANU = 11.20·SANU – 6.56). Taken together, these results support the hypothesis that proximal arm non-use is driven mainly by elbow extension non-use, while shoulder antepulsion non-use plays a less consistent role.

- What are the limits of your analysis (by ChatGPT)

- The sample size was limited, which reduces statistical power and may exaggerate or obscure true relationships.
- Correlation and linear regression do not establish causality; the observed associations may be influenced by other compensatory strategies or biomechanical factors.
- The analysis was based on a specific reaching task under controlled conditions, which may not generalize to all functional activities.
- Measurement errors (marker placement, motion capture noise, anthropometric estimations) may affect the precision of non-use scores.
- Inter-individual variability (stroke severity, recovery stage, motor strategy) was not controlled, which could contribute to unexplained variance.

# 2.1 Statistical Concepts

## 1. What is a median? A mean?
- **Mean:** the sum of all values divided by the number of values. It is sensitive to extreme values (outliers).  
- **Median:** the middle value when the data are ordered. Less sensitive to outliers.  
*(Reference: Howell, 2013; Field, 2017)*

## 2. What is a variance? A standard deviation?
- **Variance:** the average of the squared differences between each value and the mean.  
- **Standard deviation (SD):** the square root of the variance; expressed in the same units as the data.  
*(Reference: Altman, 1991; Field, 2017)*

## 3. What is a normal distribution?
A bell-shaped, symmetrical distribution where mean = median = mode.  
About 68% of data lie within 1 SD of the mean, 95% within 2 SD.  
*(Reference: Altman & Bland, 1995, BMJ – open access)*

## 4. What is a statistical test?
A statistical test is a method used to decide, based on data, whether an observed effect is likely real or due to chance.  
*(Reference: Howell, 2013)*
